Reproducible, flexible and high throughput data extraction from primary literature: The \pkg{metaDigitise} \proglang{R} package

### 2017-11-25 docs/manuscript.tex
manuscript editting

% \usepackage[utf8]{inputenc},
% \usepackage{geometry} 
% \geometry{a4paper}  
% %\usepackage[parfill]{parskip} 
% \usepackage{graphicx}	
% \usepackage{amsmath}
% \usepackage{fullpage}
% \usepackage{setspace} 
% \usepackage{lineno}
% \usepackage[none]{hyphenat}
% \usepackage{listings}
% \usepackage[round]{natbib}

@@

%% - \Address{} of at least one author
%% - May contain multiple affiliations for each author
%%   (in extra lines, separated by \emph{and}\\).
%% - May contain multiple authors for the same affiliation
%%   (in the same first line, separated by comma).
\Address{
  Joel L. Pick, Shinichi Nakagawa, Daniel W.A. Noble\\
  Ecology and Evolution Research Centre\\
  School of Biological, Earth and Environmental Sciences\\
  University of New South Wales\\
  Kensington, NSW 2052\\
  Sydney, AUSTRALIA\\
  E-mail: \email{joel.l.pick@gmail.com}\\
  URL: \url{https://eeecon.uibk.ac.at/~zeileis/}
}

@@

In many different contexts, researchers need to make use of data presented in primary literature. Most notably, this includes meta-analysis, which is becoming increasingly common in many research fields. Meta-analysis uses effect size estimates and their sampling variance, taken from many studies, to understand whether particular effects are common across studies and to explain variation among these effects \citep{Naka2017, Glass1976, Koricheva2013, Borenstein2009}. Meta-analysis therefore relies foremost on data extracted from primary literature and more specifically statistics that have been reported in the text or tables of research papers. These summary statistics are also, however, frequently presented in figures and so need to be manually extracted using digitising programs. Although there are several existing tools to perform tasks like this (e.g. DataThief \citep{DataThief}, GraphClick \citep{GraphClick}, WebPlotDigitizer \citep{WebPlotDigitizer}), these tools are not designed specifically for meta-analysis. 
%% need to spell out clear points
%% 1 - dont manipulate data/allow different plot types
Firstly, they typically only provide the user with calibrated x,y coordinates from imported figures, and do not differentiate between common plot types that are use to present data. This means that a large amount of downstream data manipulation is subsequently required, that differs for different plots types. For example, data are frequently presented in mean and error plots (Figure \ref{fig:plot_type}A), for which the user wants a mean and error estimate for each group presented in the figure. With existing programs, x,y coordinates of means and errors are returned, to which the user must manually discern between mean and error coordinates and assign points to groups. The error then needs to be calculated as the deviation from the mean, and then transformed to a standard deviation, depending on the type of error presented.
%% 2- adding metadata
Secondly, digitising programs do not easily allow the integration of metadata at the time of data extraction, such as experimental group or variable names, and sample sizes. This makes the downstream calculations more laborious, as the information has to be added later, in most cases using different software.
%% 3 - in same statistical software and high throughput
Finally, existing programs do not import a set of images and allow the user to systematically work through them. Instead they require the user to manually import images one by one, and export data into individual files, that need to be imported and edited using different software. In essence, existing software does not provide an optimized research pipeline to facilitate data extraction, editing and reproducibility. These are major issues because extracting from figures can be an incredibly time-consuming process. Furthermore, although meta-analysis is an important tool in consolidating the data from many studies many of the processes involved in data extraction are opaque and difficult to reproduce, making extending studies problematic. Having a tool that facilitates reproducibility in meta-analyses will increase transparency and go a long way to resolving the reproducibility crises we are seeing in many fields \citep{peng_reproducible_2006, peng_reproducible_2011, sandve_ten_2013, Parker2016}.

@@

When \fct{metaDigitise} is run, it recognizes all the images in a directory and automatically imports them one by one, allowing the user to click and enter relevant information about a figure as they go. This expedites digitising figures by preventing users from having to constantly change directories and / or open new images. The data from a completed image is automatically saved in an .RDS file to a \code{caldat} directory, that is created within the parent directory when first executing the \fct{metaDigitise} function. These files enable re-plotting and editing of images at a later point (see below).

@@

%% where to put calibration???

@@

Here, the naming of the files placed in the folder will contain the paper number, first author and the figure number to keep data uniquely associated with figures. At first there is one figure in the folder, shown in Figure \ref{fig:Anderson}. Running \fct{metaDigitise} brings up a series of prompts for the user using a main menu that provides access to a number of its features:

\begin{CodeChunk}
\begin{CodeInput}
R> setwd("~/FiguresToExtract")
R> digitised_data <- metaDigitise(".", summary = TRUE)
\end{CodeInput}
\begin{CodeOutput}
	Do you want to...
1: Process new images
2: Import existing data
3: Edit existing data
Selection: 
\end{CodeOutput}
\end{CodeChunk}

@@

\fct{metaDigitise} then asks the user whether the figure needs to be rotated or flipped. This can be needed when box plots and mean and error plots are not orientated correctly. In some cases, older papers can give slightly off angled images which can be corrected by rotating. So, in this prompt the user has three options: \code{f} for "Flip", \code{r} for "rotate" or \code{c} for "continue.

@@

After this, \fct{metaDigitise} will ask the user to specify the plot type. Depending on the figure, the user can specify that it is a figure containing the mean and error (\code{m}), a box plot (\code
{b}), a scatter plot (\code{s}) or a histogram (\code{h}). If the user has specified \code{diff} instead of \code{same} in response to the question about whether the plot types are the same or different, this question will pop up for each plot, but will only be asked once if plots are all the same.

@@

 m: Mean and error
 b: Box plot
 s: Scatter plot 
 h: Histogram

 R> s
\end{CodeOutput}
\end{CodeChunk}

After selecting the figure type a new set of prompts will come up that will ask the user first what the y and x-axis variables are. This is useful as users can keep track of the different variables across figures and papers. Here, the user can just add this information in to the \proglang{R} console. Once complete, details on how to calibrate the x and y-axis appear, so that the relevant statistics / data can be correctly calculated. When working with a plot of mean and standard errors, the x-axis is rather useless in terms of calibration so \pkg{metaDigitise} just asks the user to calibrate the y-axis.

@@

\begin{CodeChunk}
\begin{CodeOutput}
What is the value of y1 ?
\end{CodeOutput}
\begin{CodeInput}
R> 4.5
\end{CodeInput}
\begin{CodeOutput}
What is the value of y2 ?
\end{CodeOutput}
\begin{CodeInput}
R> 7
\end{CodeInput}
\begin{CodeOutput}
What is the value of x1 ?
\end{CodeOutput}
\begin{CodeInput}
R> 2
\end{CodeInput}
\begin{CodeOutput}
What is the value of x2 ?
\end{CodeOutput}
\begin{CodeInput}
R> 4
\end{CodeInput}
\begin{CodeOutput}
Re-calibrate? (y/n) 
R> n
\end{CodeOutput}
\end{CodeChunk}

@@

Group identifier:
\end{CodeOutput}
\begin{CodeInput}
R> setosa
\end{CodeInput}
\begin{CodeOutput}
Click on points you want to add.
If you want to remove a point, or are finished with a
group, exit (see above), then follow prompts

Add points, delete points or continue? (a/d/c) 
\end{CodeOutput}
\begin{CodeInput}
R> c
\end{CodeInput}
\end{CodeChunk}

Once we are done digitising all the groups our plot will look something like Figure \ref{fig:scatter_extract}

@@

When completed \pkg{metaDigitise} will write the digitised data to the caldat directory, such that our new directory structure is as follows

@@

Users can access the \pkg{metaDigitise} object created (001\_Anderson\_1935\_Fig1) at any time using the metaDigitse function. In the \proglang{R} console, the summarised data for the digitised figure can be printed on screen or even written to a .csv file:

\begin{CodeChunk}
\begin{CodeInput}
R> digitised_data
\end{CodeInput}
\begin{CodeOutput}
                  filename    group_id         variable  mean  error error_type  n     r    sd  plot_type
001_Anderson_1935_Fig1.png      setosa  Sepal width (mm)  3.42  0.40  sd        39  0.75  0.40  scatterplot
001_Anderson_1935_Fig1.png      setosa  Sepal length (mm) 5.00  0.38  sd        39  0.75  0.38  scatterplot
001_Anderson_1935_Fig1.png  versicolor  Sepal width (mm)  2.77  0.32  sd        44  0.52  0.32  scatterplot
001_Anderson_1935_Fig1.png  versicolor  Sepal length (mm) 5.95  0.53  sd        44  0.52  0.53  scatterplot
\end{CodeOutput}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> setwd("~/FiguresToExtract")
R> digitised_data <- metaDigitise(".", summary = TRUE)
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeOutput}
Are all plot types the same? (diff/same) 
\end{CodeOutput}
\begin{CodeInput}
R> diff
\end{CodeInput}
\begin{CodeOutput}
**** NEW PLOT ****

@@

Flip, rotate or continue f/r/c 
\end{CodeOutput}
\begin{CodeInput}
R> c
\end{CodeInput}
\begin{CodeOutput}
Please specify the plot_type as either: mean and error, box plot,
scatter plot or histogram m/b/s/h: 
\end{CodeOutput}
\begin{CodeInput}
R> m
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeOutput}
What is the variable? 
\end{CodeOutput}
\begin{CodeInput}
R> Sepal length
\end{CodeInput}
\begin{CodeOutput}
On the Figure, click IN ORDER: 
      y1, y2

@@

What is the value of y1 ?
\end{CodeOutput}
\begin{CodeInput}
R> 5
\end{CodeInput}
\begin{CodeOutput}
What is the value of y2 ?
\end{CodeOutput}
\begin{CodeInput}
R> 6.5
\end{CodeInput}
\begin{CodeOutput}
Re-calibrate? (y/n) 
\end{CodeOutput}
\begin{CodeInput}
R> n
\end{CodeInput}
\begin{CodeOutput}
Enter sample sizes? y/n 
\end{CodeOutput}
\begin{CodeInput}
R> y
\end{CodeInput}
\begin{CodeOutput}
Group identifier: 
\end{CodeOutput}
\begin{CodeInput}
R> setosa
\end{CodeInput}
\begin{CodeOutput}
Group sample size: 
\end{CodeOutput}
\begin{CodeInput}
R> 50
\end{CodeInput}
\begin{CodeOutput}
Click on Error Bar, followed by the Mean

Add group, Delete group or Finish plot? a/d/f 
\end{CodeOutput}
\begin{CodeInput}
R> a
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeOutput}
data
                  filename    group_id         variable   mean  error error_type n    r    sd   plot_type
001_Anderson_1935_Fig1.png      setosa  Sepal width (mm)  3.42  0.40  sd        39  0.75  0.40  scatterplot
001_Anderson_1935_Fig1.png      setosa  Sepal length (mm) 5.00  0.38  sd        39  0.75  0.38  scatterplot
001_Anderson_1935_Fig1.png  versicolor  Sepal width (mm)  2.77  0.32  sd        44  0.52  0.32  scatterplot
001_Anderson_1935_Fig1.png  versicolor  Sepal length (mm) 5.95  0.53  sd        44  0.52  0.53  scatterplot
     002_Doe_2013_Fig1.png      setosa  Sepal length      5.00  0.11  se        50  NA    0.78  mean_error
     002_Doe_2013_Fig1.png  viriginica  Sepal length      6.59  0.18  se        50  NA    1.26  mean_error
     002_Doe_2013_Fig1.png  versicolor  Sepal length      5.94  0.14  se        50  NA    1.01  mean_error
     003_Doe_2013_Fig3.png      catana  Sepal length      4.95  0.36  sd        50  NA    0.36  histogram
\end{CodeOutput}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> setwd("~/FiguresToExtract")
R> digitised_data <- metaDigitise(".")
\end{CodeInput}
\begin{CodeOutput}
    Do you want to...
1: Process new images
2: Import existing data
3: Edit existing data
Selection: 
\end{CodeOutput}
\end{CodeChunk}

@@

\subsection{Adding sample sizes to previous Digitisations}
%Including addition of N later

\section{Interobserver Variability and Validation}
%% simulated data
%% x people doing y plots
\subsection{Interobserver variability in digitisations}
We evaluated the consistency of digitisation using \pkg{metaDigitise} by simulating a dataset with two different groups and two traits. These data were then used to construct four different plot types (scatterplot, mean and error, histogram and boxplots). Each variable was plotted twice for a given plot type (figures were modified slightly to give users a sense that they were digitising new data) generating a total of 14 figures. We randomised the order of these figures (using the same data) and provided the entire figure directory to a total of 15 indpendent digitisers. The same data was presented at least twice for each figure type to estimate between- and within- user variability (i.e., the intra-class correlation coefficient) in the data. Digitisers varied in their level of experience from those having done meta-analyses or comparative work to those not having any experience doing so. We got users to digitise all 14 figures and then send the data on these to evaluate interobserver variablity across the plots.

We analysed these data by first transformting each users mean, standard deviation and correlation coefficient generated from their digitisations to standardized mean differences (in otherwords, deviations were percentage differences from the true summary statistics). We then estimated intra-class correlation coefficients (ICC) (adjusted - accounting for covariates and unadjusted - not accounting for covariates) using linear mixed effect models with user identify as a random effect. We tested the significance of ICC estimates using liklihood ratio tests based on 1000 parametric bootstraps of the model. The overall model also included plot type, variable and group to account for possible biases across the variables and plots. We used standardised mean, standard deviation and correlation coeficients as response variables in seperate models using both \pkg{lme4} \citep{bates2015} and \pkg{MCMCglmm} \citep{hadfield2010}, estimating ICC using the \pkg{rptR} package \citep{stoffel2017} in R. If digitisations were consistent across all users then we would predict no significant between user variability in the data.

\subsection{Results}

  Across plot types we found no evidence for any inter-observer variability in digitisations for the mean (Unadjusted ICC = 0, 95 CI = 0 to 0.029, p = 1; Adjusted ICC = 0, 95 CI = 0 to 0.033, p = 1), standard deviation (Unadjusted ICC = ; Adjusted ICC = ) or correlation coefficient (Unadjusted ICC = ; Adjusted ICC = )

@@

%% discuss specifically everything it can't do, and jsutify it slightly

>>>>

@@

%% - \Address{} of at least one author
%% - May contain multiple affiliations for each author
%%   (in extra lines, separated by \emph{and}\\).
%% - May contain multiple authors for the same affiliation
%%   (in the same first line, separated by comma).
\Address{
  Joel L. Pick, Shinichi Nakagawa, Daniel W.A. Noble\\
  Ecology and Evolution Research Centre\\
  School of Biological, Earth and Environmental Sciences\\
  University of New South Wales\\
  Kensington, NSW 2052\\
  Sydney, AUSTRALIA\\
  E-mail: \email{joel.l.pick@gmail.com}\\
  URL: \url{https://github.com/daniel1noble/metaDigitise}
}

@@

In many different contexts, researchers need to make use of data presented in primary literature. Most notably, this includes meta-analysis, which is becoming increasingly common in many research fields. Meta-analysis uses effect size estimates and their sampling variance, taken from many studies, to understand whether particular effects are common across studies and to explain variation among these effects \citep{Glass1976,Borenstein2009,Koricheva2013,Naka2017}. Meta-analysis therefore relies foremost on data extracted from primary literature and more specifically statistics that have been reported in the text or tables of research papers. These summary statistics are also, however, frequently presented in figures and so need to be manually extracted using digitising programs. Although there are several existing tools to perform tasks like this (e.g. \proglang{DataThief} \citep{DataThief}, \proglang{GraphClick} \citep{GraphClick}, \proglang{WebPlotDigitizer} \citep{WebPlotDigitizer}), these tools are not designed specifically for meta-analysis. 
%% need to spell out clear points
%% 1 - dont manipulate data/allow different plot types

Firstly, they typically only provide the user with calibrated x,y coordinates from imported figures, and do not differentiate between common plot types that are use to present data. This means that a large amount of downstream data manipulation is subsequently required, that differs for different plots types. For example, data are frequently presented in mean and error plots (Figure \ref{fig:plot_type}A), for which the user wants a mean and error estimate for each group presented in the figure. With existing programs, x,y coordinates of means and errors are returned, to which the user must manually discern between mean and error coordinates and assign points to groups. The error then needs to be calculated as the deviation from the mean, and then transformed to a standard deviation, depending on the type of error presented.
%% 2- adding metadata

Secondly, digitising programs do not easily allow the integration of metadata at the time of data extraction, such as experimental group or variable names, and sample sizes. This makes the downstream calculations more laborious, as the information has to be added later, in most cases using different software.
%% 3 - in same statistical software and high throughput

Finally, existing programs do not import a set of images and allow the user to systematically work through them. Instead they require the user to manually import images one by one, and export data into individual files, that need to be imported and edited using different software. In essence, existing software does not provide an optimized research pipeline to facilitate data extraction, editing and reproducibility. These are major issues because extracting from figures can be an incredibly time-consuming process. Furthermore, although meta-analysis is an important tool in consolidating the data from many studies many of the processes involved in data extraction are opaque and difficult to reproduce, making extending studies problematic. Having a tool that facilitates reproducibility in meta-analyses will increase transparency and go a long way to resolving the reproducibility crises we are seeing in many fields \citep{peng_reproducible_2006, peng_reproducible_2011, sandve_ten_2013, Parker2016,Ihle2017}.

@@

When \fct{metaDigitise} is run, it recognizes all the images in a directory and automatically imports them one by one, allowing the user to click and enter relevant information about a figure as they go. This expedites digitising figures by preventing users from having to constantly change directories and / or open new images. The data from a completed image is automatically saved as a \code{metaDigitise} object in an .RDS file to a \code{caldat} directory, that is created within the parent directory when first executing the \fct{metaDigitise} function. These files enable re-plotting and editing of images at a later point (see below).

@@

@@

Here, the naming of the files placed in the folder will contain the paper number, first author and the figure number to keep data uniquely associated with figures. At first there is one figure in the folder, shown in Figure \ref{fig:Anderson}. Running \fct{metaDigitise} brings up a series of prompts for the user using a main menu that provides access to a number of its features (... here represents the user's project directory):

\begin{CodeChunk}
\begin{CodeInput}
R> digitised_data <- metaDigitise(".../FiguresToExtract", summary = TRUE)
\end{CodeInput}
\begin{CodeOutput}
	Do you want to...
1: Process new images
2: Import existing data
3: Edit existing data
Selection: 
\end{CodeOutput}
\end{CodeChunk}

@@

\fct{metaDigitise} then asks the user whether the figure needs to be rotated or flipped. This can be needed when box plots and mean and error plots are not orientated correctly. In some cases, older papers can give slightly off angled images which can be corrected by rotating. So, in this prompt the user has three options: \code{f} for ``Flip'', \code{r} for ``rotate'' or \code{c} for ``continue''.

@@

After this, \fct{metaDigitise} will ask the user to specify the plot type. Depending on the figure, the user can specify that it is a figure containing the mean and error (\code{m}), a box plot (\code
{b}), a scatter plot (\code{s}) or a histogram (\code{h}). If the user has specified \code{d} instead of \code{s} in response to the question about whether the plot types are the same or different, this question will pop up for each plot, but will only be asked once if plots are all the same.

@@

 m: Mean and error
 b: Box plot
 s: Scatter plot 
 h: Histogram
\end{CodeOutput}
\begin{CodeInput}
R> s
\end{CodeInput}
\end{CodeChunk}

After selecting the figure type a new set of prompts will come up that will ask the user first what the y and x-axis variables are. This is useful as users can keep track of the different variables across figures and papers. Here, the user can just add this information in to the \proglang{R} console. Once complete, details on how to calibrate the x and y-axis appear, so that the relevant statistics / data can be correctly calculated. When working with a plot of mean and standard errors, the x-axis is rather useless in terms of calibration so \fct{metaDigitise} just asks the user to calibrate the y-axis.

@@

\begin{CodeChunk}
\begin{CodeOutput}
What is the value of y1 ?
\end{CodeOutput}
\begin{CodeInput}
R> 4.5
\end{CodeInput}
\begin{CodeOutput}
What is the value of y2 ?
\end{CodeOutput}
\begin{CodeInput}
R> 7
\end{CodeInput}
\begin{CodeOutput}
What is the value of x1 ?
\end{CodeOutput}
\begin{CodeInput}
R> 2
\end{CodeInput}
\begin{CodeOutput}
What is the value of x2 ?
\end{CodeOutput}
\begin{CodeInput}
R> 4
\end{CodeInput}
\begin{CodeOutput}
Re-calibrate? (y/n) 
\end{CodeOutput}
\begin{CodeInput}
R> n
\end{CodeInput}
\end{CodeChunk}

@@

Group identifier:
\end{CodeOutput}
\begin{CodeInput}
R> setosa
\end{CodeInput}
\begin{CodeOutput}
Click on points you want to add.
If you want to remove a point, or are finished with a
group, exit (see above), then follow prompts
\end{CodeOutput}
\end{CodeChunk}

To finish selecting points, the user can exit following the instructions (e.g. by pressing ESC on a mac). The user is then asked if they want to add or delete points from that group.

\begin{CodeChunk}
\begin{CodeOutput}
Add or Delete points to this group, or Continue? (a/d/c) 
\end{CodeOutput}
\begin{CodeInput}
R> c
\end{CodeInput}
\end{CodeChunk}

Once we are done digitising all the groups our plot will look something like Figure \ref{fig:scatter_extract}.

@@

When completed \fct{metaDigitise} will write the digitised data as a \code{metaDigitise} object to a RDS file in the caldat directory, such that our new directory structure is as follows

@@

Users can access the \code{metaDigitise} object created (001\_Anderson\_1935\_Fig1) at any time using the metaDigitse function. In the \proglang{R} console, the summarised data for the digitised figure can be printed on screen or even written to a .csv file:

\begin{CodeChunk}
\begin{CodeInput}
R> digitised_data
\end{CodeInput}
{\scriptsize
\begin{CodeOutput}
                  filename    group_id         variable  mean  error error_type  n     r    sd  plot_type
001_Anderson_1935_Fig1.png      setosa  Sepal width (mm)  3.42  0.40  sd        39  0.75  0.40  scatterplot
001_Anderson_1935_Fig1.png      setosa  Sepal length (mm) 5.00  0.38  sd        39  0.75  0.38  scatterplot
001_Anderson_1935_Fig1.png  versicolor  Sepal width (mm)  2.77  0.32  sd        44  0.52  0.32  scatterplot
001_Anderson_1935_Fig1.png  versicolor  Sepal length (mm) 5.95  0.53  sd        44  0.52  0.53  scatterplot
\end{CodeOutput}
}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> digitised_data <- metaDigitise(".../FiguresToExtract", summary = TRUE)
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeOutput}
Are all plot types Different or the Same? (d/s)
\end{CodeOutput}
\begin{CodeInput}
R> d
\end{CodeInput}
\begin{CodeOutput}
**** NEW PLOT ****

@@

Flip, rotate or continue (f/r/c) 
\end{CodeOutput}
\begin{CodeInput}
R> c
\end{CodeInput}
\begin{CodeOutput}
Please specify the plot_type as either:

 m: Mean and error
 b: Box plot
 s: Scatter plot 
 h: Histogram
\end{CodeOutput}
\begin{CodeInput}
R> m
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeOutput}
What is the y variable? 
\end{CodeOutput}
\begin{CodeInput}
R> Sepal length
\end{CodeInput}
\begin{CodeOutput}
On the Figure, click IN ORDER: 
      y1, y2

@@

What is the value of y1 ?
\end{CodeOutput}
\begin{CodeInput}
R> 5
\end{CodeInput}
\begin{CodeOutput}
What is the value of y2 ?
\end{CodeOutput}
\begin{CodeInput}
R> 6.5
\end{CodeInput}
\begin{CodeOutput}
Re-calibrate? (y/n) 
\end{CodeOutput}
\begin{CodeInput}
R> n
\end{CodeInput}
\begin{CodeOutput}
Do you know sample sizes? (y/n)  
\end{CodeOutput}
\begin{CodeInput}
R> y
\end{CodeInput}
\begin{CodeOutput}
If there are multiple groups, enter unique group identifiers (otherwise press enter)
Group identifier: 
\end{CodeOutput}
\begin{CodeInput}
R> setosa
\end{CodeInput}
\begin{CodeOutput}
Group sample size: 
\end{CodeOutput}
\begin{CodeInput}
R> 50
\end{CodeInput}
\begin{CodeOutput}
Click on Error Bar, followed by the Mean

Add group, Edit Group, Delete group or Finish plot? (a/e/d/f) 
\end{CodeOutput}
\begin{CodeInput}
R> a
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeChunk}
{\scriptsize
\begin{CodeOutput}
data
                  filename    group_id         variable   mean  error error_type n    r    sd   plot_type
001_Anderson_1935_Fig1.png      setosa  Sepal width (mm)  3.42  0.40  sd        39  0.75  0.40  scatterplot
001_Anderson_1935_Fig1.png      setosa  Sepal length (mm) 5.00  0.38  sd        39  0.75  0.38  scatterplot
001_Anderson_1935_Fig1.png  versicolor  Sepal width (mm)  2.77  0.32  sd        44  0.52  0.32  scatterplot
001_Anderson_1935_Fig1.png  versicolor  Sepal length (mm) 5.95  0.53  sd        44  0.52  0.53  scatterplot
     002_Doe_2013_Fig1.png      setosa  Sepal length      5.00  0.11  se        50  NA    0.78  mean_error
     002_Doe_2013_Fig1.png  viriginica  Sepal length      6.59  0.18  se        50  NA    1.26  mean_error
     002_Doe_2013_Fig1.png  versicolor  Sepal length      5.94  0.14  se        50  NA    1.01  mean_error
     003_Doe_2013_Fig3.png      catana  Sepal length      4.95  0.36  sd        50  NA    0.36  histogram
\end{CodeOutput}
}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> digitised_data <- metaDigitise(".../FiguresToExtract")
\end{CodeInput}
\begin{CodeOutput}
    Do you want to...
1: Process new images
2: Import existing data
3: Edit existing data
Selection: 
\end{CodeOutput}
\end{CodeChunk}

@@

\subsection{Adding sample sizes to previous Digitisations}
%Including addition of N later
In many cases important information, such as sample sizes, may not be readily available or clear when digitising figures. In these circumstances users will have answered `no' to the question about whether they have sample sizes or not while digitising. To expedite finding and adding in these sample sizes to do the necessary calculations (if for example a figure presented 95\% CI's or standard errors), \fct{metaDigitise} has s specific edit option that allows users to enter in previously omitted sample sizes. It works by first identifying the missing sample sizes in the digitised output, re-plotting the relevant figure and then prompting the user to enter the sample sizes for the relevant groups in the figure, one by one. As an example, assume that we were missing sample sizes for two groups in 002\_Doe\_2013\_Fig1.png:

\begin{CodeChunk}
{\scriptsize
\begin{CodeOutput}
            filename    group_id         variable   mean  error error_type n    r    sd   plot_type
002_Doe_2013_Fig1.png      setosa  Sepal length      5.00  0.11  se        NA  NA    NA  mean_error
002_Doe_2013_Fig1.png  viriginica  Sepal length      6.59  0.18  se        NA  NA    NA  mean_error
\end{CodeOutput}
}
\end{CodeChunk}

Here, we can see that we are missing the sample sizes for setosa and viriginica, and as a result, sd is not calculated because \fct{metaDigitise} needs this information to make the calculation. If the user found this information after contacting the authors for clarification then they can add these back in as follows:

\begin{CodeChunk}
\begin{CodeInput}
R> digitised_data <- metaDigitise(".../FiguresToExtract")
\end{CodeInput}
\end{CodeChunk}

\begin{CodeChunk}
\begin{CodeOutput}
Do you want to...

1: Process new images
2: Import existing data
3: Edit existing data

Selection:
\end{CodeOutput}
\begin{CodeInput}
R> 3
\end{CodeInput}
\end{CodeChunk}

\begin{CodeChunk}
\begin{CodeOutput}
Choose how you want to edit files:

1: Cycle through images
2: Choose specific file to edit
3: Enter previously omitted sample sizes

Selection:
\end{CodeOutput}
\begin{CodeInput}
>R 3
\end{CodeInput}
\end{CodeChunk}

\fct{metaDigitise} will replot the figure after this and list, only the groups missing data, for which the user can then update the data. This is then re-integrated back into the data automatically and the sd calculated.

\begin{CodeChunk}
\begin{CodeOutput}
Group " setosa ": Enter sample size  
\end{CodeOutput}
\begin{CodeInput}
R> 50
\end{CodeInput}
\begin{CodeOutput}
Group " viriginica ": Enter sample size
\end{CodeOutput}
\begin{CodeInput}
R> 50
\end{CodeInput}
\end{CodeChunk}

\section{Inter-observer Variability and Validation}

\subsection{Interobserver variability in digitisations}
In order to evaluate the consistency of digitisation using \pkg{metaDigitise} between users, we simulated a dataset of two traits with two different groups. These data were then used to construct plots of the four different types (scatterplot, mean and error, histogram and boxplots). Each variable was plotted twice for each given plot type (figures were modified slightly to give users a sense that they were digitising new data) generating a total of 14 figures. 15 independent digitisers were provided with a directory with all 14 figures in a randomised order. Digitisers ran \pkg{metaDigistise} on their own computers, across different operating systems (including Mac, Windows and Linux). Digitisers varied in their level of experience, from people with experience of meta-analyses or comparative work to those without any science background. We asked users to digitise all 14 figures and collected the mean, standard deviation and correlation coefficient (for scatterplots) generated by \fct{metaDigitise} for every plot digitised. We transformed these data to standardized differences as

\begin{equation}
\frac{\theta - \hat{\theta}}{\hat{\theta}}
\end{equation}

where $\theta$ is the estimate value and $\hat{\theta}$ is the true value, meaning that deviations were percentage differences from the true summary statistics. The correlation coefficient deviation was not divided by the true value, as it is already on a standardised scale. This deviation can be seen as a measure of bias. The resulting data was used to assess between- and within- user variability (i.e., the intra-class correlation coefficient) in the data. This was done using linear mixed effect models with user identify as a random effect. Standardised mean, standard deviation and correlation coeficients were used as response variables in seperate models. Sampling variance for ICC estiamtes was generated based on 1000 parametric bootstraps of the model and the significance was tested using liklihood ratio tests. These models were run using the \pkg{lme4} \citep{bates2015} and \pkg{rptR} \citep{stoffel2017} packages in \proglang{R}.

If digitisations were consistent across all users then we should find no significant between user variability in the data. Indeed, across plot types we found no evidence for any inter-observer variability in digitisations for the mean (ICC = 0, 95\% CI = 0 to 0.029, p = 1), standard deviation (ICC = 0, 95\% CI = 0 to 0.033, p = 0.5) or correlation coefficient (ICC = 0.053, 95\% CI = 0 to 0.296, p = 0.377). There were was little bias in between digitised and true values, on average 1.63\% (mean = 0.02\%, sd = 4.9\%, r = -0.03\%) and overall there were only small absolute differences between digitised and true values, deviating, on average 2.18\% (mean = 0.40\%, sd = 5.81\%, r = 0.33\%) for all three summary statistics.

SD estimates from digitisations are clearly more prone to error than means or correaltion cefficients. This is due to the fact that SD estimation from the summary statistics derived from plots is more error prone, especially at small sample sizes \citep{Wan2014}. If the mean absoluate difference is calculated for each plot type, we can see that this effect is driven mainly by extraction from boxplots and histgrams (\% difference):
\begin{CodeOutput}
    boxplot   histogram  mean_error scatterplot 
     15.805       5.210       1.500       0.433 
\end{CodeOutput}

\subsection{Testing the accuracy of digitisations}
To test how accurate \pkg{metaDigitise} is at matching points to their true values, we generated four random scatterplots, each with 20 data points, and digitised these with \fct{metaDigitise}. This was done by one digitiser, as there is no detectable between user variation. Data digitised using \pkg{metaDigitise} was essentially perfectly correlated with the true simulated data for both the x-variable (Pearson's correlation; r = 0.9999915, t = 2137.4, df = 78, p-value < 0.001) and y-variable (r = 0.9999892, t = 1897.8, df = 78, p-value < 0.001).

@@

\begin{sidewaystable}[ph!]
{\centering
\begin{tabular}{lccccccc}
\hline
Function                  & metaDigitise & GraphClick$^1$ & DataThief$^2$  & DigitizeIt$^3$ & WebPlotDigitizer$^4$ & metagear$^5$ & digitize$^6$\\
\hline
Scatterplots              & \checkmark   & \checkmark & \checkmark & \checkmark & \checkmark     & \checkmark$^7$ & \checkmark \\
Mean and error plots      & \checkmark   & \checkmark & \checkmark & $\times$   & $\times$       & \checkmark$^7$ & $\times$    \\
Boxplots                  & \checkmark   & $\times$   & $\times$   & $\times$   & $\times$       & $\times$       & $\times$    \\
Histograms                & \checkmark   & $\times$   & $\times$   & $\times$   & \checkmark$^7$ & $\times$       & $\times$    \\
Graph rotation$^8$        & \checkmark   & \checkmark & \checkmark & \checkmark & \checkmark     & $\times$       & $\times$    \\
Groups                    & \checkmark   & \checkmark & $\times$   & \checkmark & \checkmark     & $\times$       & $\times$    \\
Entry of metadata         & \checkmark   & $\times$   & $\times$   & $\times$   & $\times$       & $\times$       & $\times$    \\
Summarising data          & \checkmark   & $\times$   & $\times$   & $\times$   & $\times$       & $\times$       & $\times$    \\
Multiple image processing & \checkmark   & $\times$   & $\times$   & $\times$   & $\times$       & $\times$       & $\times$    \\
Reproducable$^9$  & \checkmark   & \checkmark & \checkmark & $\times$   & \checkmark     & $\times$       & $\times$    \\
Automated point detection & $\times$     & \checkmark &     ?      & \checkmark & \checkmark     & \checkmark     & $\times$    \\
Line extraction           & $\times$     & \checkmark & \checkmark & \checkmark & \checkmark     & $\times$       & $\times$    \\
Zoom                      & $\times$     & \checkmark & \checkmark & \checkmark &       ?        & $\times$       & $\times$    \\
Log axis                  & $\times$ *   & \checkmark & \checkmark & \checkmark & \checkmark     & $\times$       & $\times$    \\
Dates                     & $\times$ *   & $\times$   & \checkmark & $\times$   & \checkmark     & $\times$       & $\times$    \\
Asymmetric error bars     & $\times$     & $\times$   & \checkmark & $\times$   & $\times$       & $\times$       & $\times$    \\
Freeware                  & \checkmark$^{10}$ & \checkmark$^{11}$ & \checkmark$^{11}$ & $\times$$^{11}$ & \checkmark$^{11}$ & \checkmark$^{10}$ & \checkmark$^{10}$\\
\hline
